---
title: "Plate Discipline"
author: "Joshua Mould"
date: "2/28/2021"
output:
  pdf_document: default
  html_document: default
---

# Quantifying MLB Hitter Plate Discipline

Import Necessary Packages
```{r setup, include=FALSE}


#install.packages("pacman")

pacman::p_load(tidyverse, modelr, baseballr, gam, randomForest, caret, xgboost, mlr, ggthemes, e1071, remotes, formatR, Boruta, stringi)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
options(scipen = 999)


GetAUC <- function(model, data, Target){
library(pacman)
p_load(ROCR)

pred <- predict(model, type="response")
  prediction <- prediction(pred, Target)
  Perf <- performance(prediction, "auc")
  return(as.numeric(Perf@y.values[1]))

}
```

Load in EAGLE RDS data directly rather than running the functions.
```{r}

data_2016 <- readRDS("EAGLE_2016.rds")
data_2017 <- readRDS("EAGLE_2017.rds")
data_2018 <- readRDS("EAGLE_2018.rds")
data_2019 <- readRDS("EAGLE_2019.rds")
data_2020 <- readRDS("EAGLE_2020.rds")
data_2021 <- readRDS("EAGLE_2021.rds")



```


Create combined season data frame with commonly used metrics and EAGLE so that we have each hitter from 2016-2021
```{r}
eye_2016 <- data_2016 %>% filter(PA >= 200, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS),  kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 

eye_2017 <- data_2017 %>% filter(PA >= 200, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS),  kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 

eye_2018 <- data_2018 %>% filter(PA >= 200, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS), kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 

eye_2019 <- data_2019 %>% filter(PA >= 200, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS), kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 

eye_2020 <- data_2020 %>% filter(PA >= 150, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS), kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 

eye_2021 <- data_2021 %>% filter(PA >= 200, !is.na(expRA_swing)) %>% group_by(batter, Name) %>% summarize(year = first(season.x), runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)), EAGLE = wRA / n(), avg = first(BA), obp = first(OBP), ops = first(OPS), kperc = first(SO.x) / first(PA), bbperc = first(BB.x) / first(PA), perc_swing_out_zone = sum(ifelse(strike_prob < .3 & swing == 1, 1, 0)) / sum(ifelse(strike_prob < .3, 1, 0))) 


eye_all <- eye_2016 %>% full_join(eye_2017, by = c("batter")) %>% full_join(eye_2018, by = c("batter")) %>% full_join(eye_2019, by = c("batter")) %>% full_join(eye_2020, by = c("batter")) %>% full_join(eye_2021, by = c("batter"))

eye_all <- eye_all %>% rename(wRL_2016 = runs_lost.x, wRL_2017 = runs_lost.y, wRL_2018 = runs_lost.x.x, wRL_2019 = runs_lost.y.y, wRL_2020 = runs_lost.x.x.x, wRL_2021 = runs_lost.y.y.y)

```


Create data frame containing player stats from consecutive years so that we can see predictabililty and correlation across years
```{r}
eye_this_next1 <- eye_2016 %>% left_join(eye_2017, by = c("batter", "Name")) 
eye_this_next2 <- eye_2017 %>% left_join(eye_2018, by = c("batter", "Name"))
eye_this_next3 <- eye_2018 %>% left_join(eye_2019, by = c("batter", "Name"))

eye_this_next <- bind_rows(eye_this_next1, eye_this_next2, eye_this_next3)

remove(eye_this_next1, eye_this_next2, eye_this_next3)

```


Look at results of outcome probabilities xgboost model
```{r}
all_years <- rbind(data_2016, data_2017, data_2018, data_2019, data_2020, data_2021)

levs = c("Miss", "Foul", "Out", "Single", "Double", "Triple", "Homerun")

all_years %>% 
  filter(!is.na(outcome)) %>% 
  group_by(outcome) %>% 
  summarize(Miss = mean(Miss, na.rm = T), 
            Foul = mean(Foul, na.rm = T), 
            Out = mean(Out, na.rm = T), 
            Single = mean(Single, na.rm = T), 
            Double = mean(Double, na.rm = T), 
            Triple = mean(Triple, na.rm = T), 
            Homerun = mean(Homerun, na.rm = T)) %>% 
  mutate(outcome =  factor(outcome, levels = levs)) %>%
  arrange(outcome)

```


Run linear regression models in order to test for significant relationships between EAGLE and other significant statistics
```{r}

####
# Test EAGLE correlation from year to year
eye_this_next_mod <- lm(EAGLE.y ~ EAGLE.x, data = eye_this_next)
summary(eye_this_next_mod)

####

# Test with AVG
ba_eye_mod <- lm(avg.x ~ EAGLE.x, data = eye_this_next)
summary(ba_eye_mod)

next_ba_eye_mod <- lm(avg.y ~ EAGLE.x, eye_this_next)
summary(next_ba_eye_mod)

both_ba_eye_mod <- lm(avg.y ~ EAGLE.x + avg.x, eye_this_next)
summary(both_ba_eye_mod)

# Test for OBP
obp_eye_mod <- lm(obp.x ~ EAGLE.x, data = eye_this_next)
summary(obp_eye_mod)

next_obp_eye_mod <- lm(obp.y ~ EAGLE.x, eye_this_next)
summary(next_obp_eye_mod)

both_obp_eye_mod <- lm(obp.y ~ EAGLE.x + obp.x, eye_this_next)
summary(both_obp_eye_mod)

# Now for OPS
ops_eye_mod <- lm(ops.x ~ EAGLE.x, data = eye_this_next)
summary(ops_eye_mod)

next_ops_eye_mod <- lm(ops.y ~ EAGLE.x, eye_this_next)
summary(next_ops_eye_mod)

ops_ops_eye_mod <- lm(ops.y ~ ops.x, eye_this_next)
summary(ops_ops_eye_mod)

both_ops_eye_mod <- lm(ops.y ~ EAGLE.x + ops.x, eye_this_next)
summary(both_ops_eye_mod)

###
```


#Create graphs for important relationships

First create graph for walk percentage while also testing the relationship with a linear regression model
```{r}
eye_this_next %>% filter(EAGLE.x >= 0) %>% ggplot(aes(x = EAGLE.x, y = bbperc.x)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and BB%", 
       subtitle = "2016-2021") +
  xlab("EAGLE") +
  ylab("BB%")

bbperc_mod <- lm(bbperc.x ~ EAGLE.x, data = eye_this_next)
summary(bbperc_mod)

```

Graph for relationship with O% and also test it with a linear regression model
```{r}
eye_this_next %>% filter(EAGLE.x >= 0 & perc_swing_out_zone.x < .8) %>% ggplot(aes(x = EAGLE.x, y = perc_swing_out_zone.x)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and O%", 
       subtitle = "2016-2021") +
  xlab("EAGLE") +
  ylab("O%")

operc_mod <- lm(perc_swing_out_zone.x ~ EAGLE.x, data = eye_this_next)
summary(operc_mod)
```


Create graph for relationship of EAGLE with OBP
```{r}
eye_this_next %>% 
  filter(!is.na(EAGLE.x), !is.na(obp.x), EAGLE.x >= .025) %>% 
  ggplot(aes(EAGLE.x, obp.x)) + 
  geom_point() + 
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  ggtitle("Correlation to OBP") + 
  xlab("EAGLE") + 
  ylab("OBP")

```


Create graph to highlight the correlation of EAGLE from year to year
```{r}
eye_this_next %>% filter(EAGLE.x >= 0) %>% ggplot(aes(x = EAGLE.x, y = EAGLE.y)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLEs from year to year", 
       subtitle = "2016-2021") +
  xlab("EAGLE 1") +
  ylab("EAGLE 2")

```

Create graphs showing the relationship of EAGLE with the current year's OPS and with the next year's OPS
```{r}
eye_this_next %>% filter(EAGLE.x >= 0) %>% ggplot(aes(x = EAGLE.x, y = ops.x)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and OPS", 
       subtitle = "2016-2021") +
  xlab("EAGLE") +
  ylab("OPS")

eye_this_next %>% filter(EAGLE.x >= 0) %>% ggplot(aes(x = EAGLE.x, y = ops.y)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and Next Year's OPS", 
       subtitle = "2016-2019") +
  xlab("EAGLE") +
  ylab("Next Year's OPS")
```


Bring the data of all the years together to calculate predictions from percentiles in addition to testing correlation. Also check out who is the best/worst in EAGLE
```{r}

all_eye <- all_years %>% 
  mutate(season.x = as.numeric(season.x)) %>% 
  filter(PA >= 300, !is.na(expRA_swing)) %>% 
  group_by(Name, season.x) %>% 
  summarize(runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), 
            wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)),
            ba = first(BA),
            obp = first(OBP),
            slg = first(SLG),
            ops = first(OPS),
            pitches = n(),
            bb_perc = first(BB.x) / first(PA),
            o_perc = sum(ifelse(strike_prob < .5 & swing == 1, 1, 0), na.rm = T) / sum(ifelse(strike_prob < .5, 1, 0), na.rm = T)) %>% 
  mutate(EAGLE = wRA / pitches)

all_eye %>% select(Name, season.x, EAGLE, bb_perc, o_perc) %>% arrange(-EAGLE) %>% head(30)
all_eye %>% select(Name, season.x, EAGLE, bb_perc, o_perc) %>% arrange(EAGLE) %>% head(10)

eagle_perc <- quantile(all_eye$EAGLE, probs = c(.1, .25, .75, .9))
percentiles <- data_frame(percentile = c("10th percentile", "25th percentile", "75th percentile", "90th percentile"), EAGLE.x = eagle_perc)
percentiles %>% add_predictions(ops_eye_mod, var = "OPS")
percentiles %>% add_predictions(next_ops_eye_mod, var = "Next_OPS")

all_eye %>% ggplot(aes(EAGLE)) + geom_density()

cor(all_eye$EAGLE, all_eye$ops)

cor(all_eye$o_perc, all_eye$ops)
```

Current Year EAGLE correlated with current year statistics and next year's EAGLE
```{r}
eye_this_next %>% 
  ungroup() %>% 
  select(EAGLE.x, avg.x, obp.x, ops.x, kperc.x, bbperc.x, perc_swing_out_zone.x, EAGLE.y) %>% 
  rename(o_perc = perc_swing_out_zone.x, EAGLE_cur_year = EAGLE.x, EAGLE_next_year = EAGLE.y) %>% 
  cor(use = "complete.obs")

```


Inverse EAGLE, look at EAGLE for pitchers. Who is the most deceptive/good at getting batters to make the wrong decision by swinging at bad pitches or taking good pitches
```{r}
all_pitchers_eye <- all_years %>% 
  mutate(season.x = as.numeric(season.x)) %>% 
  filter(IP >= 20, !is.na(expRA_swing)) %>% 
  group_by(pitcher_name, season.x) %>% 
  summarize(runs_lost = sum(runs_lost_bad_dec) / sum(abs_eRA_s), 
            wRA = sum(ifelse(swing == 1, expRA_swing, -expRA_swing)),
            IP = first(IP),
            ERA = first(ERA),
            K_perc= first(SO_perc),
            BB_perc = first(uBB_perc),
            xwOBA = mean(estimated_woba_using_speedangle, na.rm = T),
            pitches = n(),
            o_perc = sum(ifelse(strike_prob < .5 & swing == 1, 1, 0), na.rm = T) / sum(ifelse(strike_prob < .5, 1, 0), na.rm = T)) %>% 
  mutate(EAGLE = wRA / pitches)

all_pitchers_eye %>% select(pitcher_name, season.x, EAGLE, IP, ERA, K_perc, BB_perc, o_perc) %>% filter(IP >= 90 | IP >= 40 & season.x == 2020) %>% arrange(EAGLE)

all_pitchers_eye %>% select(pitcher_name, season.x, EAGLE, IP, ERA, K_perc, BB_perc, o_perc) %>% filter(IP >= 40 & IP <= 85 & season.x != 2020 | IP <= 35 & season.x == 2020) %>% arrange(EAGLE)



```


Check out the relationship of pitching EAGLE with other common pitcher metrics.
```{r}
all_pitchers_eye %>% 
  filter(IP >= 50) %>% 
  ggplot(aes(x = EAGLE, y = ERA)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between Inverse EAGLE and ERA", 
       subtitle = "2016-2021 (min. 40 IP)") +
  xlab("EAGLE") +
  ylab("ERA")


qualified_pitchers <- all_pitchers_eye %>% filter(IP >= 50)

ERA_mod <- lm(ERA ~ EAGLE, data = qualified_pitchers)
summary(ERA_mod)
```


```{r}
all_pitchers_eye %>% filter(IP >= 50) %>% ggplot(aes(x = EAGLE, y = K_perc)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and K%", 
       subtitle = "2016-2019") +
  xlab("EAGLE") +
  ylab("K%")

k_perc_mod <- lm(K_perc ~ EAGLE, data = qualified_pitchers)
summary(k_perc_mod)
```


```{r}


all_pitchers_eye%>% filter(IP >= 50) %>% ggplot(aes(x = EAGLE, y = BB_perc)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and BB%", 
       subtitle = "2016-2019") +
  xlab("EAGLE") +
  ylab("BB%")
```


```{r}
all_pitchers_eye%>% filter(IP >= 50) %>% ggplot(aes(x = EAGLE, y = o_perc)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and O%", 
       subtitle = "2016-2019") +
  xlab("EAGLE") +
  ylab("O%")

```


```{r}

all_pitchers_eye%>% filter(IP >= 50) %>% ggplot(aes(x = EAGLE, y = xwOBA)) + 
  geom_point() +
  stat_smooth(method = "lm") + 
  theme_gdocs() + 
  labs(title = "Relationship between EAGLE and xwOBA", 
       subtitle = "2016-2019") +
  xlab("EAGLE") +
  ylab("xwOBA")

```




